---
title: "Project2"
author: "Joseph Kahadze"
date: "March 27, 2019"
output: html_document
---

## SQL

```{r setupdb, include=FALSE}

library(tidyverse)
library(dplyr)
db <- DBI::dbConnect(RSQLite::SQLite(), "C:\\Users\\jkahadze\\Downloads\\lahman2016.sqlite")
```

# PROBLEM 1

To calculate the total payroll and the winning percentage I joined the Salaries and Teams tables on the teamID and the yearID and then filtered all the enteties with yearID >= 1990 or less than <= 2014. 

I used an inner join to filter out values with missing data so that only values with corresponding yearID and teamID. Furthermore, I reviewed the table myself to make sure there were no missing enteries.

```{sql p1, connection=db, eval=TRUE, output.var="payroll_df"}

SELECT t.teamID, t.franchID, t.yearID, t.W, t.L, t.G, s.sum_sal, ((t.W * 1.0)/(t.G * 1.0)*100.0) AS WPERC FROM Teams as t inner join (SELECT teamID, yearID, sum(salary) as sum_sal FROM Salaries GROUP BY teamID, yearID) as s ON t.teamID = s.teamID and t.yearID = s.yearID WHERE t.yearID >= 1990 and t.yearID <=2014

```

# PROBLEM 2

I used a boxplot to graph the distribution of payrolls across teams because it visually shows the mean and variance of the payroll per year.

```{r 1}

sum_df <- payroll_df

sum_df %>%
  ggplot(mapping=aes(x=factor(yearID), y=sum_sal)) +
    geom_boxplot() + scale_y_continuous(labels = scales::comma) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# Question 1

The mean payroll in the MLB increased from 1990 to 2014 from about 25,000,000 to about 100,000,000. Furthermore, the variation in payroll between teams also increased from 1990 to 2014.

# PROBLEM 3

I calculated the mean payroll for each year and then plotted it for every year to show that the payroll has a tendency to increase over time.

```{r 2}
mean_df <- payroll_df %>%
  group_by(yearID) %>%
  dplyr::summarise(mean_sal = mean(sum_sal))


mean_df %>%
  ggplot(mapping=aes(x=factor(yearID), y=mean_sal)) +
    geom_bar(stat="identity")  +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
    
```

# PROBLEM 4

I split the payroll_df into five categories based on yearID in 5 year ranges. Then I calculated the mean winning percentage and mean payroll for each of the 5 tables and plotted them.

```{r 3}
cut_df <- payroll_df

cut_df$group <- cut_df$yearID %>%
  cut(breaks=5) 

X <- split(cut_df, cut_df$group)

per1 <- X[[1]]
per1 <- per1 %>% group_by(teamID) %>% dplyr::summarise(mean_sal = mean(sum_sal), mean_wp = mean(WPERC))
per2 <- X[[2]]
per2 <- per2 %>% group_by(teamID) %>% dplyr::summarise(mean_sal = mean(sum_sal), mean_wp = mean(WPERC))
per3 <- X[[3]]
per3 <- per3 %>% group_by(teamID) %>% dplyr::summarise(mean_sal = mean(sum_sal), mean_wp = mean(WPERC))
per4 <- X[[4]]
per4 <- per4 %>% group_by(teamID) %>% dplyr::summarise(mean_sal = mean(sum_sal), mean_wp = mean(WPERC))
per5 <- X[[5]]
per5 <- per5 %>% group_by(teamID) %>% dplyr::summarise(mean_sal = mean(sum_sal), mean_wp = mean(WPERC))


per1 %>%
  ggplot(mapping=aes(x=mean_sal, y=mean_wp)) +
    geom_point() + scale_x_continuous(labels = scales::comma) +        ggtitle("1990-1995") + geom_smooth(method=lm)

per2 %>%
  ggplot(mapping=aes(x=mean_sal, y=mean_wp)) +
    geom_point() + scale_x_continuous(labels = scales::comma) +        ggtitle("1995-2000") + geom_smooth(method=lm)

per3 %>%
  ggplot(mapping=aes(x=mean_sal, y=mean_wp)) +
    geom_point() + scale_x_continuous(labels = scales::comma) +        ggtitle("2000-2004") + geom_smooth(method=lm)

per4 %>%
  ggplot(mapping=aes(x=mean_sal, y=mean_wp)) +
    geom_point() + scale_x_continuous(labels = scales::comma) +        ggtitle("2004-2009") + geom_smooth(method=lm)

per5 %>%
  ggplot(mapping=aes(x=mean_sal, y=mean_wp)) +
    geom_point() + scale_x_continuous(labels = scales::comma) +        ggtitle("2009-2014") + geom_smooth(method=lm)
```

# QUESTION 2

There is a positive correlation between the mean payroll and mean winning percentage for every 5 year period. The most significant positive correlation in the periods 1995 to 2000 and 2004 to 2009.

# PROBLEM 5

I calculated the standardized formula using the given formula st_payroll = (pay_roll - mean_payroll)/st_dev

```{r 4}

values <- payroll_df %>%
  group_by(yearID) %>%
  dplyr::summarise(mean_sal = mean(sum_sal), sd_sal = sd(sum_sal)) %>%
  inner_join(payroll_df, by="yearID") %>%
  mutate (z = (((sum_sal * 1.0) - (mean_sal*1.0)) / (sd_sal*1.0) ))

values
```

# PROBLEM 6 

I split the payroll_df into five categories based on yearID in 5 year ranges. Then I created a standardized variable z to standardize payroll. Finally I graphed all 5 split tables with mean standardized value on the x-axis and the mean winning percentage on the y-axis.

```{r 5}

values <- payroll_df %>%
  group_by(yearID) %>%
  dplyr::summarise(mean_sal = mean(sum_sal), sd_sal = sd(sum_sal)) %>%
  inner_join(payroll_df, by="yearID") %>%
  mutate (z = (((sum_sal * 1.0) - (mean_sal*1.0)) / (sd_sal*1.0) ))

values

scut_df <- values

scut_df$group <- scut_df$yearID %>%
  cut(breaks=5) 

X <- split(scut_df, scut_df$group)

per1 <- X[[1]]
per1 <- per1 %>% group_by(teamID) %>% dplyr::summarise(mean_z = mean(z), mean_wp = mean(WPERC))
per2 <- X[[2]]
per2 <- per2 %>% group_by(teamID) %>% dplyr::summarise(mean_z = mean(z), mean_wp = mean(WPERC))
per3 <- X[[3]]
per3 <- per3 %>% group_by(teamID) %>% dplyr::summarise(mean_z = mean(z), mean_wp = mean(WPERC))
per4 <- X[[4]]
per4 <- per4 %>% group_by(teamID) %>% dplyr::summarise(mean_z = mean(z), mean_wp = mean(WPERC))
per5 <- X[[5]]
per5 <- per5 %>% group_by(teamID) %>% dplyr::summarise(mean_z = mean(z), mean_wp = mean(WPERC))


per1 %>%
  ggplot(mapping=aes(x=mean_z, y=mean_wp)) +
    geom_point() + scale_x_continuous(labels = scales::comma) +        ggtitle("1990-1995") + geom_smooth(method=lm)

per2 %>%
  ggplot(mapping=aes(x=mean_z, y=mean_wp)) +
    geom_point() + scale_x_continuous(labels = scales::comma) +        ggtitle("1995-2000") + geom_smooth(method=lm)

per3 %>%
  ggplot(mapping=aes(x=mean_z, y=mean_wp)) +
    geom_point() + scale_x_continuous(labels = scales::comma) +        ggtitle("2000-2004") + geom_smooth(method=lm)

per4 %>%
  ggplot(mapping=aes(x=mean_z, y=mean_wp)) +
    geom_point() + scale_x_continuous(labels = scales::comma) +        ggtitle("2004-2009") + geom_smooth(method=lm)

per5 %>%
  ggplot(mapping=aes(x=mean_z, y=mean_wp)) +
    geom_point() + scale_x_continuous(labels = scales::comma) +        ggtitle("2009-2014") + geom_smooth(method=lm)
    
```

# QUESTION 3

The standardized plots follow similar trends but have a mean closer to 0. You can see parallels between specific time periods in the dots spread and patter.

# PROBLEM 7
  
To plot the overall standardized value for the entire time period I dind't split the data frame. I just calculated the stardadized value and graphed it using a scatterplot from 1994 to 2014.
  
```{r 6}
values <- payroll_df %>%
  group_by(yearID) %>%
  dplyr::summarise(mean_sal = mean(sum_sal), sd_sal = sd(sum_sal)) %>%
  inner_join(payroll_df, by="yearID") %>%
  mutate (z = (((sum_sal * 1.0) - (mean_sal*1.0)) / (sd_sal*1.0) ))

values <- values %>% group_by(teamID) %>% dplyr::summarise(mean_z = mean(z), mean_wp = mean(WPERC))

values %>%
  ggplot(mapping=aes(x=mean_z, y=mean_wp)) +
    geom_point() + scale_x_continuous(labels = scales::comma) +        ggtitle("1990-2014") + geom_smooth(method=lm)
    
```  

# PROBLEM 8

I calculated the effeciency using the expected winning percentage and the actual winning percentage based on the payroll and the plotted the data for the 5 given teams using a line graph. I plotted the efficiency on the y-axis and year on the x-axis. 

```{r 7}
values <- payroll_df %>%
  group_by(yearID) %>%
  dplyr::summarise(mean_sal = mean(sum_sal), sd_sal = sd(sum_sal)) %>%
  inner_join(payroll_df, by="yearID") %>%
  mutate (z = (((sum_sal * 1.0) - (mean_sal*1.0)) / (sd_sal*1.0) ))

values <- values %>%
  mutate(exp_wp = 50.0 + (2.5 * z)) %>%
  mutate(efficiency = WPERC - exp_wp) %>%
  filter(teamID == "OAK" | teamID == "BOS" | teamID == "NYA" | teamID == "ATL" | teamID == "TBA")

values %>%
  ggplot(aes(x=factor(yearID), y=efficiency, group=teamID)) +
  geom_smooth(aes(color=teamID), se=FALSE)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
    
``` 

# Question 4

This plot shows the effieciency of the teams from 1990 to 2014. Compared to plots 2 and 3 which showed the relationship between winning percentage and payroll, this graph shows the caluclated efficiency for specific teams over a period of time.

The graph shows a clear peak in Oaklands efficiency during the "Moneyball period," from 2000 to 2005. But, then the efficiency dips in later years. 

```{r disconnectdb, include=FALSE}
DBI::dbDisconnect(db)
```
