---
title: "Project 4: Interactive Data Visualization and Mapping"
author: "Joseph Kahadze"
date: "May 14, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(stringr)
library(leaflet)
library(webshot)

arrest_tab <- read_csv("http://www.hcbravo.org/IntroDataSci/misc/BPD_Arrests.csv")
```

My interactive map shows every instance of an arrest in the city of Baltimore from the BPD Arrests data set with each arrest represented by a circle. The circles are colored based on the district of the particular crime to highlight district boundaries and the density of arrests by district. Furthermore, I included a legend on the bottom right of the interactive map to show which district is represented by which color. 

```{r p1}
dat <- arrest_tab %>% 
  filter(!is.na(`Location 1`)) %>%
  separate(`Location 1`, c("lat","lng"), sep=",") %>%
  mutate(lat=as.numeric(str_replace(lat, "\\(", ""))) %>%
  mutate(lng=as.numeric(str_replace(lng, "\\)", ""))) %>%
  sample_n(2000)
dat
```

```{r p2, fig.width = 8.5, fig.height = 5.5}
pal <- colorFactor(c("black", "red", "yellow", "brown", "blue", "orange", "purple", "navy", "maroon"), domain = c("SOUTHERN", "NORTHEASTERN", "WESTERN", "NORTHERN", "NORTHWESTERN", "CENTRAL", "SOUTHWESTERN", "SOUTHEASTERN", "EASTERN"))

balto_map <- leaflet(dat) %>%
  addTiles() %>%
  setView(lat=39.29, lng=-76.61, zoom=11) %>%
  addCircleMarkers(
    radius = 6,
    color = ~pal(district),
    stroke = FALSE, fillOpacity = 0.3) %>%
  addLegend("bottomright", pal = pal, values = ~district,
    title = "District by Color",
    opacity = 1
  )

balto_map
```